The Design and Implementation of the Synchronous Language \CEU

### 2016-10-06 paper.tex
(>) Noemi corrections.

The semantics of \CEU implies an implementation with a runtime stack and a
sequential scheduler.
In contrast with Esterel's semantics, this prevents programs to be compiled
directly and efficiently to hardware~\cite{esterel.constructive}.
It also makes difficult to statically determine worst-case reaction times
towards hard real-time systems~\cite{esterel.fixed}.

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependencies
(a.k.a. \emph{instantaneous dialogue}~\cite{rp.book}).
%
\CEU takes a different approach, posing a restriction that an \code{await} is
only valid for the next reaction, i.e., if an \code{await} and \code{emit}
occur simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous
termination and re-execution of statements in the same reaction (known as
\emph{schizophrenic statements}~\cite{esterel.constructive}).
%
Also, this approach does not require data dependency analysis like Esterel
(e.g., \emph{``(set or reset of a signal must always precede any test of this
signal (signal coherence law and constructivity)''}~\cite{esterel.saxort}).

@@

The compilation process of programs in \CEU is composed of three main phases:
the \emph{parsing phase} converts the source code in \CEU to an abstract syntax 
tree (\emph{AST});
the \emph{concurrency checks phase} detects inconsistencies in programs, such as 
unbounded loops and suspicious concurrent statements;
the \emph{code generation phase} converts the \emph{AST} to standard \C code 
and augments it with platform-dependent functionality (e.g., system calls) and the 
runtime of \CEU, compiling everything with \emph{gcc} to generate the final 
binary.

@@

Note that, as discussed in Section~\ref{sec.ceu.checks}, the concurrency 
checks of \CEU infer precisely trails that are concurrent and yet do not 
share resources.
Hence, these non-conflicting trails could potentially execute with real 
parallelism in multiple cores.
%
However, our experiments with multi-core execution are actually slower than 
single-core execution in the same system.
Considering that we use \CEU primarily in control-dominated applications, this 
result is not surprising and also appears in related 
work~\cite{esterel.multicore.1,esterel.multicore.2}.
One reason is the overhead from continuous fork-and-rejoin in small reactions.
Another reason is contention from excessive locality of data in stackless 
trails sharing contiguous static memory.

@@

Terra is a system for programming wireless sensor network applications which 
uses \CEU as its scripting language~\cite{ceu.terra}.
%
Figure \ref{fig.terra} shows the three basic elements of Terra:
\CEU as the scripting language,
a set of customized pre-built components,
and the embedded virtual-machine engine which can disseminate and install 
bytecode images dynamically.
%
This approach aims to combine the flexibility of remotely uploading code with 
the expressiveness and safety guarantees of \CEU.

@@

\CEU has a strong influence from Esterel but differ in the most fundamental 
aspect of the notion of time.
%
In \CEU, instead of clock ticks, atomic external event occurrences that define 
time units.
%
The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
\CEU encrusts this event-triggered notion of time in the core of the language,
which is a prerequisite for the concurrency checks that enable safe
shared-memory concurrency.

In \CEU, internal events support stack-based micro reactions within external
reactions, providing more fine-grained control for intra-reaction execution.
%
Some variants of Statecharts also distinguish internal from external
events~\cite{statecharts.variants}.
%Some variants of Statecharts support local events which imply handling intra
%reactions like \CEU.
In Statemate, \emph{``reactions to external and internal events (...) can be
sensed only after completion of the step''}~\cite{statecharts.statemate},
implying queue-based execution.
In Stateflow, \emph{``the receiving state (of the event) acts here as a
function''}~\cite{statecharts.stateflow}, which is similar to \CEU's
stack-based execution.
To avoid recursion, \CEU adopts delayed awaits, while in Statemate,
\emph{``loops in the broadcasting of events (between states) are forbidden''}.

Like \CEU, many synchronous languages rely on deterministic scheduling to 
preserve intra-reaction determinism (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%
%It is the responsibility of the programmer to specify the execution order for 
%threads, based on either explicit priorities, or source code lexical order 
%(similar to \CEU).
%
In addition, \CEU also performs concurrency checks to detect trails that, when
reordered, change the observable behavior of the program, i.e., trails that
actually rely on deterministic scheduling.
%
%These languages have a tick-based notion of time similar to Esterel, which 
%prevents the concurrency checks of \CEU.
\emph{SCCharts}~\cite{sccharts.1} proposes \emph{sequentially constructive
programs} that allow \emph{``shared variables to have multiple values per tick
as long as these values are explicitly ordered by sequential
statements''}~\cite{sccharts.2}.
This relaxes Esterel's restriction for \emph{read-after-write} accesses only
(e.g., a signal must be emitted to be present) and permits
\emph{write-after-read} accesses typical in imperative programs (e.g., test and
set).
Although this approach still restricts concurrent accesses, it does not impose
strict deterministic scheduling like in \CEU.

\emph{Esterel+Delay}~\cite{esterel.delays} is a non-intrusive extension to
Esterel to program in terms of physical time with \code{delay} statements.
The global transformation relies on a special platform statement the describes
the available timers in the system (e.g., sample-driven or interrupt-driven).
It then expands \code{delay} statements into existing Esterel statements in a
way that the desired semantics can be realised in the platform.
%
\CEU makes physical time a special input event that feeds the runtime with an
associated time to elapse, which is decremented from all awaiting trails.
The compensation scheme of \CEU guarantees that trails awake in the correct
order and that errors are not propagated on subsequent awaits.
Interrupt-driven timers are also supported with a hook callback that the
runtime calls whenever the program awaits an earlier timer.

Regarding resource management, Esterel supports a finalization mechanism to 
unconditionally execute a series of statements on abortion.
%
In addition, \CEU also tracks pointers representing resources that cross \C 
boundaries and forces the programmer to provide associated finalizers.

@@

For the past three years, we have been teaching \CEU to undergraduate and 
graduate students in courses on \emph{distributed systems} and \emph{reactive 
programming}.
%, in which the students designing protocols, embedded systems, and graphical 
%applications.
%
Our experience shows that students take advantage of the sequential-imperative 
style of \CEU and can implement non-trivial concurrent applications in a few 
weeks.
%
More recently, a company specialized in embedded systems (not related to our 
research group) released a product based on \CEU.

>>>>

The semantics of \CEU implies an implementation with a runtime stack and a
sequential scheduler.
Unlike Esterel's semantics, this prevents programs to be compiled directly and
efficiently to hardware~\cite{esterel.constructive}.
It also makes difficult to statically determine worst-case reaction times
towards hard real-time systems~\cite{esterel.fixed}.

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependencies
(a.k.a. \emph{instantaneous dialogue}~\cite{rp.book}).
%
\CEU takes a different approach, posing a restriction that an \code{await} is
only valid for the next reaction, i.e., if an \code{await} and \code{emit}
occur simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous
termination and re-execution of statements in the same reaction (known as
\emph{schizophrenic statements}~\cite{esterel.constructive}).
%
Also, unlike Esterel, this approach does not require data dependency analysis
(e.g., \emph{``(set or reset of a signal must always precede any test of this
signal (signal coherence law and constructivity)''}~\cite{esterel.saxort}).

@@

The compilation process of programs in \CEU is composed of three main phases:
the \emph{parsing phase} converts the source code in \CEU to an abstract syntax 
tree (\emph{AST});
the \emph{concurrency checks phase} detects inconsistencies in programs, such
as unbounded loops and suspicious concurrent statements;
the \emph{code generation phase} converts the \emph{AST} to standard \C code 
and augments it with platform-dependent functionality (e.g., system calls) and
the runtime of \CEU, compiling everything with \emph{gcc} to generate the final 
binary.
%
An alternative implementation generates bytecode for an embedded virtual
machine, which can disseminate and install programs dynamically.

@@

Note that, as discussed in Section~\ref{sec.ceu.checks}, the concurrency 
checks of \CEU detect trails that are concurrent and yet do not share
resources.
Hence, these non-conflicting trails could potentially execute with real 
parallelism in multiple cores.
%
However, our experiments with multi-core execution are actually slower than 
single-core execution on the same system.
Considering that we use \CEU primarily in control-dominated applications, this 
result is not surprising and also appears in related 
work~\cite{esterel.multicore.1,esterel.multicore.2}.
One reason is the overhead from continuous fork-and-rejoin in small reactions.
Another reason is contention from excessive locality of data in stackless 
trails sharing contiguous static memory.

@@

Terra is a system for programming wireless sensor network applications which 
uses \CEU as its scripting language~\cite{ceu.terra}.
%
This approach aims to combine the flexibility of remotely uploading code with 
the expressiveness and safety guarantees of \CEU.
%
Figure \ref{fig.terra} shows the three basic elements of Terra:
\CEU as the scripting language,
a set of customized pre-built components,
and the embedded virtual-machine engine which can disseminate and install 
bytecode images dynamically.

@@

\CEU was strongly influenced by Esterel but they differ in the most fundamental 
aspect of the notion of time.
%
In \CEU, instead of clock ticks, atomic external event occurrences define time
units.
%
The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
\CEU encrusts this event-triggered notion of time in the core of the language,
and is a prerequisite for the concurrency checks that enable safe shared-memory
concurrency.

In \CEU, internal events support stack-based micro reactions within external
reactions, providing more fine-grained control for intra-reaction execution.
%
Some variants of the Statecharts synchronous visual language also distinguish
internal from external events~\cite{statecharts.variants}.
%Some variants of Statecharts support local events which imply handling intra
%reactions like \CEU.
In Statemate~\cite{statecharts.statemate}, \emph{``reactions to external and
internal events (...) can be sensed only after completion of the step''},
implying queue-based execution.
In Stateflow~\cite{statecharts.stateflow}, \emph{``the receiving state (of the
event) acts here as a function''}, which is similar to \CEU's
stack-based execution.
To avoid recursion, \CEU adopts delayed awaits, while in Statemate,
\emph{``loops in the broadcasting of events (between states) are forbidden''}.

Like other synchronous languages
(%
    \emph{Reactive C}~\cite{rp.rc},                 % 1991
    \emph{Protothreads}~\cite{wsn.protothreads},    % 2006
    \emph{SOL}~\cite{wsn.sol},                      % 2007
    \emph{SC}~\cite{rp.synchc},                     % 2009
    and
    \emph{PRET-C}~\cite{rp.pretc}%                  % 2010
),
\CEU also relies on deterministic scheduling to preserve intra-reaction
determinism.
%
%It is the responsibility of the programmer to specify the execution order for 
%threads, based on either explicit priorities, or source code lexical order 
%(similar to \CEU).
%
In addition, it also performs concurrency checks to detect trails that, when
reordered, change the observable behavior of the program, i.e., trails that
actually rely on deterministic scheduling.
%
%These languages have a tick-based notion of time similar to Esterel, which 
%prevents the concurrency checks of \CEU.
The \emph{SCCharts} formalism~\cite{sccharts.1} proposes \emph{sequentially
constructive programs} that allow \emph{``shared variables to have multiple
values per tick as long as these values are explicitly ordered by sequential
statements''}~\cite{sccharts.2}.
This relaxes Esterel's restriction for \emph{read-after-write} accesses only
(e.g., a signal must be emitted to be present) and permits
\emph{write-after-read} accesses typical in imperative programs (e.g., test and
set).
This approach still restricts concurrent accesses but does not impose
strict deterministic scheduling as \CEU does.

\emph{Esterel+Delay}~\cite{esterel.delays} is a non-intrusive extension to
Esterel to program in terms of physical time with \code{delay} statements.
The global transformation relies on a special platform statement that describes
available timers in the system (e.g., sample-driven or interrupt-driven).
It then expands \code{delay} statements into existing Esterel statements in a
way that the desired semantics can be realised in the platform.
%
In contrast, \CEU makes physical time a special input event that feeds the
runtime with an associated time to elapse, which is decremented from all
awaiting trails.
The compensation scheme of \CEU guarantees that trails awake in the correct
order and that errors are not propagated on subsequent awaits.
Interrupt-driven timers are also supported with a hook callback that the
runtime calls whenever the program awaits on an earlier timer.

Regarding resource management, Esterel supports a finalization mechanism to 
unconditionally execute a series of statements on abortion.
%
As a complement to finalization, \CEU also tracks pointers representing
resources that cross \C boundaries and forces the programmer to provide
associated finalizers.

@@

For the past three years, we have been teaching \CEU to undergraduate and 
graduate students in courses on \emph{distributed systems} and \emph{reactive 
programming}.
%, in which the students designing protocols, embedded systems, and graphical 
%applications.
%
Our experience shows that students take advantage of the sequential-imperative 
style of \CEU and can implement non-trivial concurrent applications in a few 
weeks.
%
More recently, a company specialized in embedded systems (not related to our 
research group) has released a product based on \CEU.
